\MakeUppercase{Layered Partitions of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-05-23 k-planar.tex
Added a couple more applications

\begin{document}
\maketitle

\begin{abstract}
  Layered partitions of graphs (with small layered width, in which the quotient has small treewidth) are a recently-introduced tool that have been used to solve longstanding problems on queue layouts, non-repetitive colouring, and 3-d graph drawing.  Such partitions are known to exist for planar graphs and, more generally, apex-minor-free graphs.  In the current paper, we prove that every $k$-planar graph has a layered partition of layered width $O(k^2)$ in  which the quotient graph has treewidth $O(k^3)$. This is the first result of this type for a non-minor-closed class of graphs and implies that $k$-planar graphs have both queue number and non-repetitive chromatic number upper-bounded by a function of $k$. The former result was previously shown by combining layered partitions of planar graphs with \textit{ad hoc} methods. The latter result is new, for all $k\ge 1$.  These results extend to $(g,k)$-planar graphs, the natural generalization of $k$-planar graphs to to genus-$g$ surfaces (rather than the genus-$0$ plane).
\end{abstract}

\section{Introduction}

@@

@@

\subsection{Non-Repetitive Colouring}

@@

\begin{cor}
  For every $k$-planar graph $G$, $\pi(G)\le 12(2k+3)(k+1)\times 4^{(1/6)(k+2)(k+3)(k+4)}$.
\end{cor}

@@

\subsection{Centered Colourings}

@@

\begin{cor}
  For every $(g,1)$-planar graph $G$, $g\ge 1$, $\chi_p(G)\le O(gp^4)$.
\end{cor}

@@

@@

\begin{enumerate}
  \item clustered colouring;
  \item connection with queue layouts;
  \item $d$-map graphs, which are $d^2$-planar;
  \item connection with low-treewidth/treedepth colourings;
  \item An important open question is whether $k$-planar graphs have layered $H$-partitions whose layered width is upper bounded by a function of $k$ but in which the treewidth of $H$ is bounded by an absolute constant, independent of $k$.
  \item Are our current $H$-partitions $K_{3,s}$-free for some $s=s(k)$?  If so, this is good news for clustered colouring.
\end{enumerate}

>>>>

\begin{document}
\begin{titlepage}
\maketitle

\begin{abstract}
  Layered partitions of graphs (with small layered width, in which the quotient has small treewidth) are a recently-introduced tool that have been used to solve longstanding problems on queue layouts, non-repetitive colouring, and 3-d graph drawing.  Such partitions are known to exist for planar graphs and, more generally, apex-minor-free graphs.  In the current paper, we prove that every $k$-planar graph has a layered partition of layered width $O(k^2)$ in  which the quotient graph has treewidth $O(k^3)$. This is the first result of this type for a non-minor-closed class of graphs and implies that $k$-planar graphs have both queue number and non-repetitive chromatic number upper-bounded by a function of $k$. The former result was previously shown by combining layered partitions of planar graphs with \textit{ad hoc} methods. The latter result is new, for all $k\ge 1$.  These results extend to $(g,k)$-planar graphs, the natural generalization of $k$-planar graphs to to genus-$g$ surfaces (rather than the genus-$0$ plane).
\end{abstract}
\end{titlepage}

\tableofcontents
\newpage
\section{Introduction}

@@

Here we discuss some of the consequences of Theorems~\ref{thm:1-planar}--\ref{thm:g-k-planar} for other graph parameters that can be bounded using layered partitions and other graph classes that are $k$-planar.

\subsection{Applications to Colouring}

@@

\subsubsection{Non-Repetitive Colouring}

@@

\begin{cor}\corlabel{first}
  For every $k$-planar graph $G$, $\pi(G)\le 12(2k+3)(k+1)\times 4^{(1/6)(k+2)(k+3)(k+4)}$.
\end{cor}

@@

\subsubsection{Centered Colourings}

@@

\begin{cor}\corlabel{last}
  For every $(g,1)$-planar graph $G$, $g\ge 1$, $\chi_p(G)\le O(gp^4)$.
\end{cor}

@@

\subsection{Other Graph Classes}

We conclude by pointing out two other graph classes, namely $(g,d)$-map graphs and $(g,\delta)$-string graphs that are $(g,k)$-planar for $k=O(d^2)$ and $k=\delta^2$, respectively.  This implies that Theorems~\ref{thm:1-planar}--\ref{thm:g-k-planar} and Corollaries~\ref{cor:first}--\ref{cor:last} apply to $(g,d)$-map graphs and to $(g,\delta)$-string graphs.

\subsubsection{Map Graphs}

Map graphs are defined as follows. Start with a graph $G_0$ embedded in a surface of Euler genus $g$, with each face labelled a `nation' or a `lake', where each vertex of $G_0$ is incident with at most $d$ nations. Let $G$ be the graph whose vertices are the nations of $G_0$, where two vertices are adjacent in $G$ if the corresponding faces in $G_0$ share a vertex. Then $G$ is called a \emph{$(g,d)$-map graph}. 
% A $(0,d)$-map graph is called a (plane) \emph{$d$-map graph}; such graphs have been extensively studied \citep{FLS-SODA12,DFHT05,CGP02,Chen07,Chen01}. The $(g,3)$-map graphs are precisely the graphs of Euler genus at most $g$ (see \citep{DEW17}). So $(g,d)$-map graphs provide a natural generalisation of graphs embedded in a surface.

The following lemma comes from the natural drawing of a map graph obtained by positioning each vertex of $G$ inside the corresponding nation and each edge of $G$ as a curve passing through the corresponding vertex of $G_0$. See \citet[Page~4]{dujmovic.eppstein.ea:structure}.

\begin{lem}
  Every $(g, d)$-map graph is $(g, \floor{\frac{d-2}{2}}\ceil{\frac{d-2}{2}})$-planar.
\end{lem}

\subsubsection{String Graphs}

For a surface $S$ and a graph $G$, an \emph{$S$-string representation} $f$ of $G$ is obtained by representing each vertex of $v\in V(G)$ as a curve $f(v)$ in $S$ so that, for any pair of vertices $v,w\in V(G)$, the curves $f(v)$ and $f(w)$ intersect if and only if $vw\in E(G)$.  If $G$ has maximum degree $\delta$, then an $S$-representation of $G$ is also called an $(S,\delta)$-string representation.  A graph $G$ is a \emph{$(g,\delta)$-string graph} if it has an $(S,\delta)$-string representation where $S$ is a surface of Euler genus $g$.

\begin{lem}
  Every $(g,\delta)$-string graph $G$ is a $(g,\delta^2)$-planar graph.
\end{lem}

\begin{proof}
  Consider some $(S,\delta)$-string representation, $f$, of $G$.  We will describe an embedding of $G$ into $S$ that is an $(S,k)$-plane graph.  Each vertex $v\in v(G)$ is embedded at $f(v)(0)$. For each edge $vw\in E(G)$ there is an intersection point $p$ between $f(v)$ and $f(w)$, i.e., $f(v)(t_0)=f(w)(t_1)=p$ for some $t_0,t_1\in[0,1]$.  Embed the edge $vw$ very close to the curve whose image is $\{f(v)(t): 0\le t\le t_1\}\cup f(w)(t): 0 \le t\le t_2\}$. In this embedding, for every vertex $x\in V(G)$, every portion of the curve $f(x)$ between two crossing points has up to $\delta$ edges of $G$ embedded close to it. The embedding of any edge $vw\in E(G)$ crosses at most $\delta$ other curves $f(x_1),\ldots,f(x_d)$ and each of these curves $f(x_i)$ is used to embed at most $\delta$ edges incident to $x_i$.  Therefore the embedding of any edge $vw\in E(G)$ crosses at most $\delta^2$ other edges.  
\end{proof}

@@

\begin{enumerate}
  % \item clustered colouring;
  \item connection with queue layouts;
  % \item $d$-map graphs, which are $d^2$-planar;
  \item connection with low-treewidth/treedepth colourings;
  \item An important open question is whether $k$-planar graphs have layered $H$-partitions whose layered width is upper bounded by a function of $k$ but in which the treewidth of $H$ is bounded by an absolute constant, independent of $k$.
  \item Are our current $H$-partitions $K_{3,s}$-free for some $s=s(k)$?  If so, this is good news for clustered colouring.
\end{enumerate}
